Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-10-11 paper/paper.tex

In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Readers familiar with kernel density estimation may not be familiar with some of the nuances of manifolds. Therefore, before introducing our own estimator we first introduce kernel density estimation for data in Euclidean space, then illustrate in Section \ref{Pellet} how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold. In Section \ref{MetLearn} we describe the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq} we derive our own novel distortion corrected kernel density estimate in Section \ref{DCKDE}. This allows for kernel density estimation on unknown manifolds using any popular dimension reduction algorithm. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers less familiar with differential geometry in Appendix \ref{riemgeo}.

In the following we denote as \(M\) the \(d\)-dimensional manifold from which our data are sampled. Points on this manifold are denoted \(\pmb{p}\) in general with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample. Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m>>d\). However this need not be that case, for instance \(\pmb{p}_i\) may be probability distributions on a statistical manifold. The methods we propose for estimating the density at \(\pmb{p}_i\) only require that there be available some sense of `input' distance \(d(\pmb{p}_i,\pmb{p}_j)\). We will apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\). Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(||\pmb{y}||)=K(0).
\end{align}
Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation. For instance condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise. In our empirical section more general kernel functions can be, and are, employed.

@@

To extend this to the case where data are on a manifold \(\pmb{p}_i\in\mathcal{M}\), for \(i=1,\dots,N\), \textcite{Pelletier2005-vu} propose the following estimator,

@@

The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore before providing formal details we will briefly discuss the intuition behind including this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\). However the volume of a ball of radius \(r\) around \(\pmb{p}\) must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center. The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.

@@

To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case there the manifold is not known, but where an output embedding from a dimension reduction algorithm is available requires additional estimates of the Riemannian metric. Formally, the Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\). The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).}, can be used to define geometric quantities for example angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems. This is precisely the situation when data on a manifold are mapped to \(d\)-dimensional Euclidean vectors \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) via a dimension reduction algorithm. After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion. To alleviate this issue \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\) at each data point. These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)\).

@@

>>>>

In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in Section \ref{Pellet} how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold. In Section \ref{MetLearn} we describe the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq} we derive our own novel distortion corrected kernel density estimate in Section \ref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in Appendix \ref{riemgeo}.

In the following we denote as \(M\) the \(d\)-dimensional manifold from which our data are sampled. Points on this manifold are denoted \(\pmb{p}\) in general with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample. Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m>>d\). However this need not be that case, for instance the \(\pmb{p}_i\) may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each \(\pmb{p}_i\) only require that some sense of distance between the `input' points \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\). Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(||\pmb{y}||)=K(0).
\end{align}
Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation. For instance condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise. In our empirical section more general kernel functions can be, and are, employed.

@@

For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,

@@

The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore before providing formal details we will briefly discuss the intuition behind including this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\). However the volume of this ball must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center. The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.

@@

To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where an output embedding from a dimension reduction algorithm is available, requires an estimate of the Riemannian metric. Formally, the Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\). The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\), can be used to define geometric quantities. For example angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems. This is precisely the situation when data on a manifold are mapped to \(d\)-dimensional Euclidean vectors \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) via a dimension reduction algorithm. After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion. To alleviate this issue \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\) at each data point. These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).

@@

\caption{\label{tab:fourdhdrtable}Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
\midrule
>99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
90 & \cellcolor[HTML]{E6E6E6}{0.919} & \textbf{0.926} & \cellcolor[HTML]{E6E6E6}{\textbf{0.921}} & 0.900 & \cellcolor[HTML]{E6E6E6}{\textbf{0.798}} & 0.440 & \cellcolor[HTML]{E6E6E6}{\textbf{0.779}} & 0.428\\
50 & \cellcolor[HTML]{E6E6E6}{0.825} & \textbf{0.851} & \cellcolor[HTML]{E6E6E6}{0.831} & \textbf{0.848} & \cellcolor[HTML]{E6E6E6}{\textbf{0.659}} & 0.512 & \cellcolor[HTML]{E6E6E6}{\textbf{0.623}} & 0.448\\
5 & \cellcolor[HTML]{E6E6E6}{0.508} & \textbf{0.556} & \cellcolor[HTML]{E6E6E6}{0.514} & \textbf{0.562} & \cellcolor[HTML]{E6E6E6}{\textbf{0.431}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.220}} & 0.000\\
\addlinespace
1 & \cellcolor[HTML]{E6E6E6}{0.080} & \textbf{0.190} & \cellcolor[HTML]{E6E6E6}{0.090} & \textbf{0.230} & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000\\
\bottomrule
\end{tabular}}
\end{table}
